The extended light profiles of the Sculptor and Ursa Minor dwarf galaxies: Tides or nature?

### 2025-09-09 thesis/methods.tex
thesis updates

To explore the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables. The present-day position,
distance modulus, LOS velocity, and proper motions are each sampled from
normal distributions given the reported uncertainties in
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
each sampled position/velocity back in time for 10 Gyr using \agama{}
\citep{agama}. Dynamical friction is not expected to impact orbits
substantially because of the low masses and large pericentres of the
dwarfs.

@@

To convert observed positions and velocities to Galactocentric
coordinates, we use the Astropy v4 Galactocentric frame
\citep{astropycollaboration+2022}. This frame assumes the Galactic
centre is at position \(\alpha = {\rm 17h\,45m\,37.224s}\),
\(\delta = -28^\circ\,56'\,10.23''\) with proper motions
\(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
\(\mu_\delta=-5.547\pm0.026 \masyr\) \citep[from the appendix and Table
2 of][]{reid+brunthaler2004}. The Galactic centre is at a distance from
the Sun of \(8.122\pm0.033\,\)kpc with a radial velocity =
\(11 + 1.9 \pm 3\,\kms\) \citep{gravitycollaboration+2018}. The Sun is
assumed to be \(20.8\pm0.3\,\)pc above the disk
\citep{bennett+bovy2019}. Using the procedure outlined in
\citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
rest frame is then
\(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
uncertainties in the reference frame are typically smaller than the
uncertainties on a dwarf galaxy's position and velocity.

@@

We adopt the Milky Way potential described in \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
each component and the total circular velocity profile for our fiducial
profile. The potential includes a stellar bulge, a thin and thick disk,
and a dark matter NFW halo.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
\caption[Circular velocity of potential]{Circular velocity profile of
\citet{EP2020} potential. The total circular velocity (thick black line)
is composed of an NFW halo (green dashed line), a think and thick
\citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
\citet{hernquist1990} bulge (blue dotted
line).}\label{fig:v_circ_potential}
\end{figure}

\subsection{Orbits of Sculptor}\label{orbits-of-sculptor}

Sculptor's orbital history is relatively well-constrained.
Fig.~\ref{fig:scl_orbits} illustrates point particle orbits for 100
samples of Sculptor's observed kinematics integrated backwards 5 Gyr in
both Galactocentric coordinate slices (\(x\), \(y\), \(z\)) and in
Galactocentric radius with time. All sampled orbits of Sculptor have
nearly the same morphology---the orbit primarily resides in the
\(y\)--\(z\) plane and has a similar number of periods and
pericentre/apocentre.

To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
smallest pericentre among all possible orbits. We achieve this by taking
the median parameters of all orbits with a pericentre less than the
0.0027th quantile pericentre, yielding a pericentre of 43 kpc. Given the
current observations, it is unlikely that Sculptor has a significantly
smaller pericentre than our selected orbit.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
potential in Galactocentric \(x\), \(y\), and \(z\) coordinates (top)
and in Galactocentric radius \(r\) versus time (bottom). The Milky Way
is at the centre with the disk lying in the \(x\)--\(y\) plane. Our
selected \texttt{smallperi} orbit is plotted in black and light blue
transparent orbits represent the past 5Gyr orbits sampled over Sculptor
observables in Table~\ref{tbl:scl_obs_props}. The orbit of sculptor is
well-constrained in this potential and it is unlikely to achieve a
smaller pericentre than our selected orbit.}\label{fig:scl_orbits}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor. The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}. The Smallperi represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects. }
\label{tbl:scl_orbits}
\begin{tabular}{lll}
\toprule
Property & Mean & SmallPeri\\
\midrule
distance / kpc & 83.2 & 82.6\\
$\pmra / \masyr$ & 0.099 & 0.134\\
$\pmdec / \masyr$ & -0.160 & -0.198\\
LOS velocity / $\kms$ & 111.2 & 111.2\\
$t_i / \Gyr$ & -8.74 & -9.43\\
$\vec{x}_{i} / \kpc$ & [16.13, 92.47, 39.63] & [-2.49, -42.78, 86.10]\\
$\vec{v}_i / \kms$ & [-2.37, -54.70, 128.96] & [-20.56, -114.83, -57.29]\\
pericentre / kpc & 53 & 43\\
apocentre / kpc & 102 & 96\\
$t_{\rm last\ peri} / {\rm Gyr}$ & -0.45 & -0.46\\
Number of pericentres & 6 & 6\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Orbits of Ursa Minor}\label{orbits-of-ursa-minor}

Similar to Sculptor, Ursa Minor has a well-constrained orbit.
Fig.~\ref{fig:umi_orbits} shows 100 random point-orbits of Ursa Minor.
Initially, we select an orbit with approximately the \(3\sigma\)
smallest pericentre as in Sculptor.

Ursa Minor's N-body orbit diverges from the point particle orbit. To
ensure the final conditions of the N-body simulation are close to the
intended final position, we iteratively adjust Ursa Minor's initial
conditions. Initially, starting with low-resolution runs, we adjust the
cylindrical actions of the initial orbit by the final difference in
actions at the end of orbital evolution. After the initial actions have
converged (2 iterations), we change the initial action angles by the
final difference in action angles. This method converges within 4
iterations to an orbit agreeing with the observed kinematics of Ursa
Minor.

\begin{table*}[t]
\centering
\caption[Ursa Minor Selected Orbits]{Properties of selected orbits for Ursa Minor. The "smallperi" orbit is the initial point orbit and the "smallperi.5" is the initial orbit for the N-body simulation. }
\label{tbl:umi_orbits}
\begin{tabular}{llll}
\toprule
Property & Mean & SmallPeri.1 & SmallPeri.5\\
\midrule
distance / kpc & $70.1$ & 64.6 & \\
$\pmra / \masyr$ & $-0.124 \pm 0.17$ & -0.158 & \\
$\pmdec / \masyr$ & $0.078\pm0.17$ & 0.05 & \\
$v_{\rm LOS} / \kms$ & $-245.9\pm 1$ & -245.75 & \\
$t_i / \Gyr$ & -8.74 & -9.53 & -9.53\\
$\vec{x}_{i} / \kpc$ & [4.88, -65.11, 50.78] & [-16.48 69.92 21.05] & [17.40, 74.51, 21.34]\\
$\vec{v}_i / \kms$ & [-34.28, 77.11, 101.38] & [16.32, 39.86, -116.99] & [14.27, 48.62, -114.08]\\
pericentre / kpc & 37 & 29 & 28\\
apocentre / kpc & 83 & 75 & 72\\
$t_{\rm last\ peri} / \Gyr$ & $-0.96$ & $-0.80$ & $-0.81$\\
Number of pericentres & 9 & 6 & 6\\
\bottomrule
\end{tabular}
\end{table*}

@@

We use \agama{} \citep{agama} to generate the initial N-body dark matter
halo. We assume galaxies are described by an NFW dark matter potential
(Eq.~\ref{eq:nfw}). We also assume the stars do not contribute to the
potential. The dark matter density is truncated in the outer regions by
\begin{equation}{
\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t),
}\end{equation} where we adopt \(r_t = 20 r_s\).

@@

From the observed properties of Sculptor and Ursa Minor, we infer
reasonable \LCDM{} initial halo conditions.

Table~\ref{tbl:derived_props} reports our inferred halo and kinematic
properties of Sculptor and Ursa Minor. First, taking the absolute
magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
\citet{woo+courteau+dekel2008} (with \(\sim\) 0.17 dex uncertainty), the
total current stellar mass of Sculptor and Ursa Minor are
\(M_\star \sim 3.1 \times 10^6 \Mo\) and
\(M_\star \sim 7 \times 10^5 \Mo\). Based on the stellar mass-\(\vmax\)
relation \citep[from][]{fattahi+2018}, see also \ref{fig:smhm}, Sculptor
and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\) and
\(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
\(z=0\) mass-concentration relation, this constraint translates into a
\(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
galaxy.

Fig.~\ref{fig:scl_halos} illustrates these estimates visually for both
galaxies. The stellar-mass \(\vmax\) constraint translates to an
estimate of \(\vmax\) only (horizontal band). The mass-concentration
relation describes the relationship between \(\vmax\) and \(\rmax\)
(diagonal linear band). And finally, we include a curved line which
illustrates the halo \(\vmax\) which has a specified initial LOS
velocity dispersion given \(\rmax\)
\(, \vcirc(R_h) / \sqrt{3} \approx \sigma_v\).
\(R_h \approx 0.24\,\kpc\) for both galaxies.

While there is some range in the choice of initial halo, reasonable
changes to the initial halo do not substantially affect the tidal
evolution for either galaxy. The observed velocity dispersion is
directly related to the mass within a half-light radius
\citep[e.g.,][]{wolf+2010}. As a result, halos with the same velocity
dispersion may differ in total mass but should have similar mass within
\(R_h\). So, the tidal effects on stars should be similar for halos with
similar velocity dispersions.

Our selected halos (in Table~\ref{tbl:initial_halos}) for each
simulation run are based on the cosmological constraints and the need to
match the present-day velocity dispersion at the end of the simulation.

@@

@@

To ensure that the initial conditions of the simulation are dynamically
relaxed and well-converged, we run a halo first in isolation using
\gadget{}. Since gravity is scale-free, we use the same isolation run
for all halos. We adopt \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\)
for the isolation halo based on Sculptor's mean properties. We run this
model for 5 Gyr (about three times the free fall timescale
\(t_{\rm ff} = \vcirc / r \approx 1.5\,\Gyr\) at \(r_{200}=36\,\)kpc).

For our simulation parameters, we adopt a softening length of
\begin{equation}\protect\phantomsection\label{eq:softening_length}{
h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
}\end{equation} See Appendix Section~\ref{sec:extra_convergence} for a
discussion of this choice, which is similar to the \citet{power+2003}
suggested softening. We use the relative tree opening criterion with the
accuracy parameter set to 0.005, and adaptive time stepping with
integration accuracy set to 0.01.

@@

Fig.~\ref{fig:numerical_convergence} illustrates how well our numerical
setup is able to reproduce the desired initial conditions, before and
after running the model in isolation. This figure shows that our
numerical methods are able to approximate well an NFW halo down to an
innermost radius that strongly depends on resolution. The larger the
number of particles, the smaller the radius that is effectively
``resolved'' in a given simulation. For the Sculptor halo shown in this
figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\) ), a
simulation with \(10^7\) particles is needed to resolve the innermost
100 pc. For reference, the half-light radius of Sculptor is roughly 100
pc, which means that at least 10 million particles would be needed to
follow faithfully its tidal evolution. Vertical arrows in
Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
defined by \citet[eq.\textasciitilde13]{power+2003} for NFW halos formed
in cosmological N-body simulations. This radius marks the region where
collisional effects driven by the finite number of particles used to
describe the innermost regions of a halo become important. The softening
length (from Eq.~\ref{eq:softening_length}) is typically a few times
smaller than the converged length.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
\caption[Numerical halo convergence]{Numerical convergence test for
circular velocity as a function of log radius for simulations with
different total numbers of particles in isolation. Residuals in bottom
panel are relative to NFW. The initial conditions are dotted, the
converged radius is marked by arrows \citep[from][ eq. 13]{power+2003},
and the softening length is marked by a vertical bar. Note that a slight
reduction in density starting around \(r = 30\,\kpc\) is expected given
our truncation choice.}\label{fig:numerical_convergence}
\end{figure}

@@

Next, we evolve the halo in the Galactic potential. We scale the relaxed
snapshot and softening length to match the initial halo in
\ref{tbl:initial_halos}, and shift the snapshot to the initial
conditions from the orbital analysis (see
Tables~\ref{tbl:scl_orbits}, \ref{tbl:umi_orbits}). We then evolve the
full N-body NFW model forward in time in the Galactic potential and
follow it in time until the present time, when the halo is closest to
the observed position of the galaxy.

\subsection{Tidal mass losses}\label{sec:shrinking_spheres}

To accurately follow the evolution of a halo, it is important to
determine the centre of the halo at each time chosen for analysis. We
use a shrinking-spheres centre method inspired by \citet{power+2003}.
First, we start with an initial centre estimate from the last timestep.
Then, we calculate the radius of all particles from the centre, remove
particles with a radius beyond the 0.975 quantile of the centre, and
recalculating the centre of mass. The procedure is repeated until the
selection radius is less than \textasciitilde1kpc or fewer than 0.1\% of
particles remain. After a centre has been chosen, we remove all unbound
particles based on the \gadget{} calculated potential of the halo. For
all future timesteps, we consider only particles retained from the
previous iteration.

@@

\subsection{The stellar component}\label{sec:painting_stars}

We ``paint'' stars onto dark matter particles using the particle-tagging
method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
symmetry. We initially assume exponential stars with
\(R_s = 0.10\,,\kpc\) for both galaxies.

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
know \(f({\cal E})\), the distribution function (phase-space density in
energy), then we assign a stellar weight to a given particle with energy
\({\cal E}\) using \begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation} While \(f({\cal E})\) is a phase-space density, the
differential energy distribution includes an additional \(g({\cal E})\)
occupation term \citep{BT1987}. We use Eddington inversion to find the
distribution function, \citep[eq. 4-140b in][]{BT1987} \begin{equation}{
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
}\end{equation} In practice the right, boundary term is zero as
\(\Psi \to 0\) as \(r\to\infty\), and if \(\rho \propto r^{-n}\) at
large \(r\) and \(\Psi \sim r^{-1}\) then
\(d\rho / d\Psi \sim r^{-n+1}\) which goes to zero provided that
\(n > 1\). We take \(\Psi\) from the underlying assumed analytic dark
matter potential. \(\rho_\star\) can be calculated from the surface
density, \(\Sigma_\star\), via the inverse Abel transform.

We find the stellar profiles created in this manner are stable in the
isolated systems and show excellent agreement with the assumed stellar
density profile.

>>>>

To explore the possible orbits of a dwarf galaxy, we perform a Monte
Carlo sampling of the present-day observables. The present-day position,
distance modulus, LOS velocity, and proper motions are each sampled from
normal distributions given the reported uncertainties in
Tables~\ref{tbl:scl_obs_props}, \ref{tbl:umi_obs_props}. We integrate
each sampled position/velocity back in time for 10 Gyr using \agama{}
\citep{agama}. Dynamical friction is not expected to impact orbits
substantially because of the low masses and large pericentres of the
dwarfs, so we assume a single point-mass particle for the backwards
integration.

@@

To convert observed positions and velocities to Galactocentric
coordinates, we use the Astropy v4 Galactocentric frame
\citep{astropycollaboration+2022}. This frame assumes the Galactic
centre is at position \(\alpha = {\rm 17h\,45m\,37.224s}\),
\(\delta = -28^\circ\,56'\,10.23''\) with proper motions
\(\mu_{\alpha*}=-3.151\pm0.018\ \masyr\) ,
\(\mu_\delta=-5.547\pm0.026 \masyr\) \citep[from the appendix and Table
2 of][]{reid+brunthaler2004}. The Galactic centre is at a distance from
the Sun of \(8.122\pm0.033\,\)kpc with a radial velocity =
\(11 + 1.9 \pm 3\,\kms\) \citep{gravitycollaboration+2018}. The Sun is
assumed to be \(20.8\pm0.3\,\)pc above the disk
\citep{bennett+bovy2019}. Using the procedure outlined in
\citet{drimmel+poggio2018}, the Solar velocity relative to the Galactic
rest frame is then in Cartesian coordinates
\(\V_\odot = [-12.9 \pm 3.0, 245.6 \pm 1.4, 7.78 \pm 0.08]\) km/s. The
uncertainties in this reference frame are typically smaller than the
uncertainties on a dwarf galaxy's position and velocity.

@@

We adopt the Milky Way potential described in \citet{EP2020}, which is
an analytic approximation to that proposed by \citet{mcmillan2011}.
Fig.~\ref{fig:v_circ_potential} plots the circular velocity profiles of
each component and the total circular velocity profile for this
potential. The potential includes a stellar bulge, a thin and thick
disk, and a dark matter NFW halo.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/v_circ_potential.pdf}}
\caption[Circular velocity of potential]{Circular velocity profile of
\citet{EP2020} potential. The total circular velocity (thick black line)
is composed of an NFW halo (green dashed line), a think and thick
\citet{miyamoto+nagai1975} disk (orange dash-dotted line), and a
\citet{hernquist1990} bulge (blue dotted line). \textbf{make radius
lowercase}}\label{fig:v_circ_potential}
\end{figure}

\subsection{Sculptor's Orbit}\label{sculptors-orbit}

Sculptor's orbital in the assumed potential is relatively
well-constrained. Fig.~\ref{fig:scl_orbits} illustrates point particle
orbits for 100 samples of Sculptor's observed kinematics integrated
backwards 5 Gyr in both Galactocentric coordinate slices (\(x\), \(y\),
\(z\)) and in Galactocentric radius with time. These orbits sample the
uncertainties in distance, proper motion, and radial velocity, as given
in Table~\ref{tbl:scl_obs_props}. All sampled orbits of Sculptor have
nearly the same morphology---the orbit primarily resides in the
\(y\)--\(z\) plane and has a similar number of periods and
pericentre/apocentre.

To maximize tidal effects, we select an orbit with the \(\sim 3\sigma\)
smallest pericentre among all possible orbits. We achieve this by taking
the median parameters of all orbits with a pericentre less than the
0.0027th quantile pericentre, yielding a pericentre of 43 kpc. Given the
current observations, it is unlikely that Sculptor has had a
significantly smaller pericentre than our selected orbit, which we refer
to as the \smallperi{} orbit.

@@

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{/Users/daniel/thesis/figures/scl_xyzr_orbits.pdf}}
\caption[Sculptor Orbits]{The orbits of Sculptor in a static Milky Way
potential in Galactocentric \(x\), \(y\), and \(z\) coordinates (top)
and in Galactocentric radius \(r\) versus time (bottom). The Milky Way
is at the centre with the disk lying in the \(x\)--\(y\) plane. Our
selected \smallperi{} orbit is plotted in black and light green
transparent orbits represent the past 5Gyr orbits sampled over Sculptor
observables in Table~\ref{tbl:scl_obs_props}. The orbit of Sculptor is
well-constrained in this potential and it is unlikely to achieve a
smaller pericentre than the \smallperi{} orbit.}\label{fig:scl_orbits}
\end{figure}

@@

\begin{table*}[t]
\centering
\caption[Sculptor Selected Orbits]{Properties of selected orbits for Sculptor. The mean orbit represents the observational mean from Table \ref{tbl:scl_obs_props}. The \smallperi{} represents instead the $3\sigma$ smallest pericentre, which we use to provide an upper limit on tidal effects. }
\label{tbl:scl_orbits}
\begin{tabular}{lll}
\toprule
Property & mean & \smallperi{}\\
\midrule
distance / kpc & 83.2 & 82.6\\
$\pmra / \masyr$ & 0.099 & 0.134\\
$\pmdec / \masyr$ & -0.160 & -0.198\\
LOS velocity / $\kms$ & 111.2 & 111.2\\
$t_i / \Gyr$ & -8.74 & -9.43\\
${x}_{i} / \kpc$ & 16.13 & -2.49\\
${y}_{i} / \kpc$ & 92.47 & -42.78\\
${z}_{i} / \kpc$ & 39.63 & 86.10\\
$\V_{x\,i} / \kms$ & -2.37, & -20.56\\
$\V_{y\,i} / \kms$ & -54.70 & -114.83\\
$\V_{z\,i} / \kms$ & 128.96 & -57.29\\
pericentre / kpc & 53 & 43\\
apocentre / kpc & 102 & 96\\
$t_{\rm last\ peri} / {\rm Gyr}$ & -0.45 & -0.46\\
Number of pericentres & 6 & 6\\
\bottomrule
\end{tabular}
\end{table*}

\subsection{Ursa Minor's Orbit}\label{ursa-minors-orbit}

Similar to Sculptor, Ursa Minor has a well-constrained orbit in the
assumed MW potential. Fig.~\ref{fig:umi_orbits} shows 100 random
point-orbits of Ursa Minor. As for Sculptor, we select an orbit with
approximately the \(3\sigma\) smallest pericentre, the Ursa Minor
\(\smallperi{}\) orbit.

We shall see later that the orbit of Ursa Minor's N-body model differs
from the point particle orbit because of the effects of tidal mass loss.
To ensure the final conditions of the N-body simulation are close to the
intended final position, we iteratively adjust Ursa Minor's initial
conditions. Initially, starting with low-resolution runs, we adjust the
cylindrical actions of the initial orbit by the final difference in
actions at the end of orbital evolution. After the initial actions have
converged (2 iterations), we change the initial action angles by the
final difference in action angles. This method converges within 4
iterations to an orbit agreeing with the observed kinematics of Ursa
Minor. Since Sculptor's orbit is less strongly affected by tides, we do
not carry out this correction for Sculptor.

\begin{table*}[t]
\centering
\caption[Ursa Minor Selected Orbits]{Properties of selected orbits for Ursa Minor. The \smallperi{} orbit is the initial point orbit and the \smallperi{}.5 is the initial orbit for the N-body simulation. }
\label{tbl:umi_orbits}
\begin{tabular}{llll}
\toprule
Property & Mean & \smallperi{}.1 & \smallperi{}.5\\
\midrule
distance / kpc & $70.1$ & 64.6 & \\
$\pmra / \masyr$ & $-0.124 \pm 0.17$ & -0.158 & \\
$\pmdec / \masyr$ & $0.078\pm0.17$ & 0.05 & \\
$v_{\rm LOS} / \kms$ & $-245.9\pm 1$ & -245.75 & \\
$t_i / \Gyr$ & -8.74 & -9.53 & -9.53\\
${x}_{i} / \kpc$ & 4.88 & -16.48 & 17.40\\
${y}_{i} / \kpc$ & -65.11 & 69.92 & 74.51\\
${z}_{i} / \kpc$ & 50.78 & 21.05 & 21.34\\
$\V_{x\,i} / \kms$ & -34.28 & 16.32 & 14.27\\
$\V_{y\,i} / \kms$ & 77.11 & 39.86 & 48.62,\\
$\V_{z\,i} / \kms$ & 101.38 & -116.99 & -114.08\\
pericentre / kpc & 37 & 29 & 28\\
apocentre / kpc & 83 & 75 & 72\\
$t_{\rm last\ peri} / \Gyr$ & $-0.96$ & $-0.80$ & $-0.81$\\
Number of pericentres & 9 & 6 & 6\\
\bottomrule
\end{tabular}
\end{table*}

@@

We use \agama{} \citep{agama} to generate the initial N-body dark matter
halo. We assume galaxies are described by a spherical, isotropic NFW
dark matter potential (Eq.~\ref{eq:nfw}). We also assume the stars do
not contribute to the potential. The dark matter density is truncated in
the outer regions by \begin{equation}{
\rho_{\rm tNFW} = e^{-(r/r_t)^3}\ \rho_{\rm NFW}(t),
}\end{equation} where we adopt \(r_t = 20 r_s\).

@@

The observed half-light radius and velocity dispersion of Sculptor and
Ursa Minor, together with the mass-concentration relation of \LCDM{}
halos, determines the properties of the N-body models adopted for each
dwarf.

Table~\ref{tbl:derived_props} lists the our inferred halo and kinematic
properties of Sculptor and Ursa Minor. First, taking the absolute
magnitudes from \citet{munoz+2018} with the mass-to-light ratio from
\citet{woo+courteau+dekel2008} (with \(\sim\) 0.17 dex uncertainty), the
total current stellar mass of Sculptor and Ursa Minor are
\(M_\star \sim 3.1 \times 10^6 \Mo\) and
\(M_\star \sim 7 \times 10^5 \Mo\), respectively. Based on the stellar
mass-\(\vmax\) relation {[}fig.~1 from \citet{fattahi+2018};{]},
Sculptor and Ursa Minor's halos should have \(\vmax \approx 31 \,\kms\)
and \(\vmax \approx 27\,\kms\). Finally, using the \citet{ludlow+2016}
\(z=0\) mass-concentration relation, this constraint translates into a
\(\rmax \approx 6 {\rm kpc}\) and \(\rmax \approx 5\,\kpc\) for each
galaxy. These choices set the velocity dispersion of a (massless)
stellar system embedded in such a halo, depending on its half-light
radius.

The situation is illustrated in Fig.~\ref{fig:scl_halos}. Here we show,
in the NFW \(\rmax\)--\(\vmax\) plane, the velocity dispersion of
stellar systems with half-light radius matching that of Sculptor and
Ursa Minor (\(R_h=240\,\)pc for both galaxies). For Scl, assuming
\(\vmax = 31\,\kms\) and \(\rmax = 6\,\kpc\) gives
\(\sigma_v \approx = XXX\), in contrast with the observed dispersion of
\textbf{add}. To match the observed value, we choose a somewhat smaller
value for r\_max=?? kpc (shown by the blue circle in the left panel of
Fig. 3.4). We have also experimented with a smaller rmax and Vmax for
Scl (yellow square in the same panel, to account for the fact that the
half-light radius and velocity dispersion may be affected by tidal mass
losses.

For Ursa Minor, the cosmological mean predicts \ldots. We chose a halo
with a higher \(\vmax\) and \(\rmax\) so the initial dispersion is
\textbf{add}. While above the observed value, we will show in the next
section this results in a final velocity dispersion near to the observed
values. \textbf{add}

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/initial_velocity.pdf}}
\caption{Initial halo velocity profiles}
\end{figure}

While there is some range in the choice of initial halo, reasonable
changes to the initial halo do not substantially affect the tidal
evolution for either galaxy. The observed velocity dispersion is
directly related to the mass within a half-light radius
\citep[e.g.,][]{wolf+2010}. As a result, halos with the same velocity
dispersion may differ in total mass but should have similar mass within
\(R_h\). So, the tidal effects on stars should be similar for halos with
similar velocity dispersions. \textbf{add curve for density at
pericentre}

Our selected halos (in Table~\ref{tbl:initial_halos}) for each
simulation run are based on the cosmological constraints and the need to
match the present-day velocity dispersion at the end of the simulation.
\textbf{Make figure of initial halos, mark observed half-light radius
and velocity dispersion, pericentric MW density?}

@@

\subsection{\texorpdfstring{The N-body code:
\gadget{}}{The N-body code: }}\label{the-n-body-code}

@@

To ensure that the initial conditions of the simulation are dynamically
relaxed and well-converged, we run a halo first in isolation using
\gadget{}. Since gravity is scale-free, we use the same isolation run
for all halos and rescale the results to the desired values of size and
mass. We adopt a fiducial value of \(\rmax = 6.0\,\)kpc and
\(\vmax = 31\,\kms\) for the isolation halo based on Sculptor's mean
properties. We run this model for 5 Gyr (about three times the crossing
time \(t_{\rm cross} = \vcirc / r \approx 1.5\,\Gyr\) at
\(r_{200}=36\,\)kpc).

For our simulation parameters, we adopt a softening length of
\begin{equation}\protect\phantomsection\label{eq:softening_length}{
h_{\rm grav} = 0.014\,{\rm kpc}\left(\frac{r_{\rm max}}{6.0\,{\rm kpc}}\right)\left(\frac{N}{10^7}\right)^{-1/2}.
}\end{equation} See Appendix Section~\ref{sec:extra_convergence} for a
discussion of this choice, which is similar to the \citet{power+2003}
suggested softening. We use in \gadget{} the relative tree opening
criterion with the accuracy parameter set to \(\alpha =0.005\) (so nodes
are opened when \(M\,l/r^3 < \alpha |a|\) for a distance from the node
of mass \(M\) of \(r\), \(l\) is the side length of the node, and \(a\)
is the particle's total acceleration), and adaptive time stepping with
integration accuracy set to \(\eta=0.01\) (particles must take
time-steps smaller than \(dt < \sqrt{2\,\eta\,h_{\rm grav} / a}\) for
acceleration \(a\)).

@@

Fig.~\ref{fig:numerical_convergence} illustrates how well our numerical
setup is able to reproduce the desired initial conditions, before and
after running the model in isolation. This figure shows that our
numerical methods are able to approximate well an NFW halo down to a
given innermost radius that strongly depends on resolution. The larger
the number of particles, the smaller the radius that is effectively
``resolved'' in a given simulation. For the Sculptor halo shown in this
figure (with \(\rmax = 6.0\,\)kpc and \(\vmax = 31\,\kms\) ), a
simulation with \(10^7\) particles is needed to resolve the innermost
100 pc. For reference, the half-light radius of Sculptor is roughly 100
pc, which means that at least 10 million particles are needed to follow
faithfully its tidal evolution. Vertical arrows in
Fig.~\ref{fig:numerical_convergence} indicate the ``convergence radius''
defined by \citet[eq.\textasciitilde13]{power+2003} for NFW halos formed
in cosmological N-body simulations. This radius marks the region where
collisional effects driven by the finite number of particles used to
describe the innermost regions of a halo become important. The softening
length (from Eq.~\ref{eq:softening_length}) is typically a few times
smaller than the converged length.

\begin{figure}
\centering
\pandocbounded{\includegraphics[keepaspectratio]{figures/iso_converg_num.pdf}}
\caption[Numerical halo convergence]{Numerical convergence test for
circular velocity as a function of log radius for simulations with
different total numbers of particles in isolation. Residuals in bottom
panel are relative to NFW. The initial conditions are dotted, the
converged radius is marked by arrows \citep[from][ eq. 13]{power+2003},
and the softening length is marked by a vertical bar. Note that a slight
reduction in density starting around \(r = 30\,\kpc\) is expected given
our outer truncation choice.}\label{fig:numerical_convergence}
\end{figure}

@@

Next, we evolve the halo in the Galactic potential. We scale the relaxed
snapshot and softening length to match the initial halo (see
Table~\ref{tbl:initial_halos}), and shift the snapshot to the initial
conditions inferred from the orbital analysis (see
Tables~\ref{tbl:scl_orbits}, \ref{tbl:umi_orbits}). We then evolve the
full N-body NFW model forward in time in the Galactic potential until
the present time, when the halo is closest to its present-day observed
position in the MW halo.

\subsection{Halo centring}\label{sec:shrinking_spheres}

To accurately follow the evolution of a halo, it is important to
determine the centre of the self-bound halo remnant at each time chosen
for analysis. We use a shrinking-spheres centre method inspired by
\citet{power+2003}. First, we start with an initial centre estimate from
the last timestep. Then, we calculate the radius of all particles from
the centre, remove particles with a radius beyond the 0.975 quantile of
the centre, and recalculate the centre of mass. The procedure is
repeated until the selection radius is less than \textasciitilde1kpc or
fewer than 0.1\% of particles remain. After a centre has been chosen, we
remove all unbound particles based on the \gadget{} calculated potential
of the halo. For all future timesteps, we consider only particles
retained from the previous iteration.

@@

\subsection{Sculptor and Ursa Minor's initial stellar
components}\label{sec:painting_stars}

We ``paint'' stars onto dark matter particles using the particle-tagging
method \citep[e.g.,][]{bullock+johnston2005}, assuming spherical
symmetry. We initially assume stars follow a projected exponential law
(Eq.~\ref{eq:exponential_law}) with \(R_s = 0.10\,,\kpc\) for both
galaxies. The tagging method assigns a probability to each dark matter
particle, which is proportional to the ``light-to-mass'' ratio required
to match the assumed stellar light profile. We briefly describe the
procedure next. We refer interested readers to \textbf{REF (rapha?)} for
further details. \emph{code may be found at?}

Let \(\Psi\) be the potential (normalized to vanish at infinity) and
\({\cal E}\) the binding energy \({\cal E} = \Psi - 1/2 v^2\). If we
know \(f({\cal E})\), the distribution function (phase-space density as
a function of energy), then we assign a stellar weight to a given
particle with energy \({\cal E}\) using \begin{equation}{
P_\star({\cal E}) = \frac{f_\star({\cal E})}{f_{\rm halo}({\cal E})}.
}\end{equation} While \(f({\cal E})\) is a phase-space density, the
differential energy distribution includes an additional \(g({\cal E})\)
occupation term \citep{BT1987}. We use Eddington inversion to find the
distribution function, \citep[eq. 4-140b in][]{BT1987} \begin{equation}{
f({\cal E}) = \frac{1}{\sqrt{8}\, \pi^2}\left( \int_0^{\cal E} \frac{d^2\rho}{d\Psi^2} \frac{1}{\sqrt{{\cal E} - \Psi}}\ d\Psi + \frac{1}{\sqrt{\cal E}} \left(\frac{d\rho}{d\Psi}\right)_{\Psi=0} \right).
}\end{equation} In practice the right, boundary term is zero as
\(\Psi \to 0\) as \(r\to\infty\), and if \(\rho \propto r^{-n}\) at
large \(r\) and \(\Psi \sim r^{-1}\) then
\(d\rho / d\Psi \sim r^{-n+1}\) which goes to zero provided that
\(n > 1\). We take \(\Psi\) from the underlying assumed analytic dark
matter potential. \(\rho_\star\) can be calculated from the surface
density, \(\Sigma_\star\), via the inverse Abel transform.

We find the stellar profiles created in this manner are stable in the
isolated systems and show excellent agreement with the assumed stellar
density profile. \textbf{Add figure showing stellar and dark matter
density profiles\ldots{}}
